A Graph Theory Based Approach to MDAO Problem Formulation

### 2013-04-02 graph_syntax.tex
updates to the intro with brian's stuff

The driver node and the implicit edge are only allowed to be pressent in PSGs. All 
other node and edge types can be present in any of the three graphs, subject to a 
few restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\item If a variable node has any incomming explicit edges, it may not have any 
  incomming implicit edges. 
\end{enumerate}

@@

Design variables, objectives, and constraints all require external decisions 
to be made about a given system to identify them. Consider 
a notional problem given by the graph in Fig. \ref{f:design vars} There are three input variables 
which, could potentially become design variables: $w$, $x$, and $y$. Because the output, $x$, of $A$
is connected to the input, $x$, of $B$ x can't actually be a design variable. The inputs $w$ and $y$
have no incomming edges and could each be selected as design variables.

\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Notional graph with two potential design variables \label{f:design vars}}
\end{figure}

@@

Objective and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
athrimetic opperations to form the final objective function. For example, the problem from 
Fig. \ref{f:design vars} might have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality of an equality, e.g. 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer explictly identifies variables 
or groups of variables and assigns a simple, explicit relationship between them.

>>>>

The driver node and the implicit edge are only allowed to be pressent in PSGs. All 
other node and edge types can be present in any of the three graphs, subject to a 
few restrictions: 
\begin{enumerate}
\item A model node can only have one edge directed to or from another model node.
\item A model node can only have fixed edges directed in or out.
\item A model node must have at least one edge directed in and at least one edge 
  directed out.
\item If a variable node has an outgoing edge to a model node then it may not have 
  any other outgoing edges.
\end{enumerate}

@@

Design variables, objectives, and constraints all require external decisions 
to be made about a given system to identify them. Consider 
a notional problem given by the graph in Fig. \ref{f:designvars} There are three input variables 
which, could potentially become design variables: $w$, $x$, and $y$. Because the output, $x$, of $A$
is connected to the input, $x$, of $B$ x can't actually be a design variable. The inputs $w$ and $y$
have no incomming edges and could each be selected as design variables.

\begin{figure}[htb!]
  \begin{center}
    \includegraphics[width=.6\textwidth]{images/design_vars_graph}
  \end{center}
  \caption{Notional graph with two potential design variables \label{f:designvars}}
\end{figure}

@@

Objective and constraints are also constructs that need to be identified by engineers 
as part of the design process. In the case of objective functions single output values 
could be selected, but commonly multiple values are assembeled together via simple 
athrimetic opperations to form the final objective function. For example, the problem from 
Fig. \ref{f:designvars} might have an objective function of $z^2$ or $2y + z$. Constraints are 
always given in the form of either an inequality of an equality, e.g. 
$\frac{2y}{10}-1<0$ or $3x+2y - 5=0$. In both cases, an engineer explictly identifies variables 
or groups of variables and assigns a simple, explicit relationship between them.

### 2013-04-02 intro.tex
updates to the intro with brian's stuff

    As the size and complexity of engineering systems grow, the time and expense for setting up 
    analysis models grow with them. Multidisciplinary Design Analysis and Optimization (MDAO)
    frameworks such as OpenMDAO\cite{Gray2012} and ModelCenter have enabled a new level of analysis tool integration 
    and paved the way for models with more analyses and increasing numbers of interdisciplinary couplings. That 
    new capability has created a new challenge since configuring a model with the larger nubmer of analyses, associated inputs and outputs
    can be very difficult. Even for models with 10's of analyses, there could be hundreds of variables that need to be managed. 
    It is not hard to imagine that the task of combining all the analyses into a consistent system model capable of solving a relevant 
    engineering design problem could easily become much more costly than createing the discipline analyses themselves. Thus, for a large system
    as the couplings between the disciplines begin to dominate the design space, the couplings between the analyses begin to dominate 
    the job of setting up the model.

    When building a model, coupling represents the reciprical flow of information between two analyses. 
    We propose the use of graph-based syntax to describe the flow of information from inputs, 
    through analyses, to outputs. If you allow for any free inputs to become design
    variables and some set of the outputs to become design objectives and constraints, then the graph 
    represents the problem formulation for a given design effort. 
    The graph-based syntax presented has application in all phases of the design process. 
    It can be used early on, to help construct a valid problem formulation from the set of available analysis tools. 
    It also provides a consistent structure to describe problem formulation that provides a foundation to apply
    algorithmic methods for selecting and then implementing effective solution strategies inside an MDAO framework.

    We've already identifed that a problem formulation graph should include design variables, analysis tools and their 
    associated inputs and outputs, and objectives and constraints. Notably absent from that list are any kind of 
    solvers, optimizers, or other iterative solution finding tools. At its most fundamental state, a problem formulation 
    includes only information about what is being sought after in a design or what the
    goals of a given problem are. It need not contain any information about solution paths or strategies to reach those goals. 
    While solution information can be represented with the graph syntax presented here, a unique aspect of this syntax is that 
    such information always remains seperable from the underlying problem formulation. In other words, a given a graph representing 
    a specific solution strategy for a specific problem formulation, it is possible to remove the solution information from the 
    graph and reclaim the more basic problem formulation.

@@

    The FPF in Eq. \ref{eqn:simple_fpf} differs from both Eq. \ref{eqn:simple} and Eq. \ref{eqn:simple2} because it has 
    two constraints which both must be met. The presence of both of these constraints fully decouples the problem so that 
    either $A$ or $B$ could be run first or both could be run simultaneously. By removing either constraint and replacing 
    it with a direct dependence between the two analyses you could regain the earlier two problem formulations. Alexandrov
    and Lewis demonstrated the value of a more modular approach to problem formulation because it enables one to transition between different 
    MDAO solution strategies depending on the specifics of the problem\cite{Alex2000}.

\section{Existing Graph--Based Syntax}

    As shown above, the mathematical language for specifying problem formulations is very general and can be used both for 
    fundamental and specific problem formulations. Tedford and Martins used the above mathematical syntax to specify the 
    FPF for a set of test problems and also to describe specific formulations for solving them with a 
    number of optimization architectures\cite{Tedford2009}. Their work demonstrates clearly how multiple specific 
    problem formulations can all relate back to a common FPF. The challenge with using this 
    traditional mathematical syntax is that it is not easily manipulated or analyzed. 
    A number of graph-based methods have been used successfully to translate the 
    mathematical syntax into a more useful computational form.

@@

    Lamb and Martins included objectives and constraint functions as nodes in an Extended 
    DSM (XDSM) in order to capture a more complete description of solution strategies for MDAO problems\cite{Lambe2012}. 
    XDSM retains the square adjacency matrix form from DSM, but by adding in the 
    new elements they partially combined a traditional DSM with an FDT. 
    This allows XDSM to represent data dependency between multiple analysis tools as well as between analysis tools and
    objective/constraint functions. With the additional information included in an 
    XDSM, Lu and Martins applied both ordering and partitioning 
    algorithms on an MDAO test problem named the Scalable Problem \cite{Lu2012}.

@@

    Of the three methods, XDSM comes the closest to fully describing a problem formulation. 
    Thought it does not contain specific variable information on the diaganol of the DSM, this could be easily added back in
    simply by putting the variables back into the graph and drawing the appropriate connections
    between them and the relevant analysis nodes. However, it's dependence on 
    solution strategy specifics to represent couplings can not be removed without 
    stripping it of the useful information contained in an FDT. We propose
    a new graph syntax that retains the features of XDSM, without the dependence on 
    solution strategy (although it does allow for that information to be present). This 
    new syntax allows for the compelete specification of an MDAO problem.

>>>>

    The number of analysis tools required in multidisciplinary design optimization (MDO) studies is growing
    in parallel with the increasing scope of typical MDO problems. An example of this growth in scope can
    be observed in the historical evolution of the disciplines involved in MDO problems in aircraft design.
    Multidisciplinary optimization emerged as a separate field from structural optimization through the
    need to introduce formal techniques for managing the coupling of aerodynamic loads and structural
    deformations, through the linking of aerodynamic vortex lattice or panel methods with structural finite
    element models [[REF]]. Subsequently, flight performance and life cycle economics tools were integrated
    into MDO analysis workflows for conceptual and preliminary design studies [[REF]]. Currently, MDO
    problems for aircraft design also often include tools for aircraft noise and emissions. In sum, it is now
    commonplace for 5-10 analysis tools to be employed for typical aircraft design optimization studies [[REF]].

    The number of analysis tools is expected to grow in the future as the scope of MDO problems continues
    to evolve, and as computing is increasingly commoditized. This expansion of scope will be driven, in
    part, by consideration of additional disciplines. Current trends on the horizon for aircraft MDO studies
    include incorporation of manufacturing analyses [[REF]], subsystem performance [[REF]], and fleet-level
    aggregate models of emissions, noise, or economics [[REF]].

    As the size and complexity of engineering systems grow, the time and expense for setting up 
    analysis models grow with them. Multidisciplinary Design Analysis and Optimization (MDAO)
    frameworks such as OpenMDAO\cite{Gray2012} and ModelCenter have enabled a new level of analysis tool integration 
    and paved the way for models with more analyses and increasing numbers of interdisciplinary couplings. That 
    new capability has created a new challenge. Even for models with 10's of analyses there could be hundreds or thousands
    of variables that need to be linked together. Commonly different analyses will provide competing values for the 
    same physical quantity, and these conflicts need to be resolved somehow. To make matters worse, the competeing values 
    may not even be in the same format. These occurrences are particularly acute when analysis tools have differing fidelities. For
    example, an abstracted aerodynamic analysis such as an empirical drag buildup model may return only
    integrated drag, whereas a CFD tool may return pressure and shear stress distributions across the entire
    surface grid. If an analysis downstream of the aerodynamics tool needs only integrated drag as an input,
    then the designer has a free choice of which of the two possible aerodynamics analysis tools to select to
    provide the drag estimate (presuming that drag can be computed from surface distributions by a simple
    integration algorithm). On the other hand, if a downstream analysis needs a pressure distribution in
    order to compute pitching moment, for example, then any feasible MDO problem formulation must
    include CFD or similar analysis in the data flow, regardless of whether the empirical drag buildup model
    is also included.

    With all the added complexity from larger models it is not hard to imagine that the task of combining all the analyses into a 
    consistent system model capable of solving a relevant engineering design 
    problem could easily become much more costly than createing the discipline 
    analyses themselves. Just as for a large system the couplings between the 
    disciplines begin to dominate the design space, the couplings between the 
    analyses begin to dominate the job of setting up the model. It is this problem of 
    determining sets of analysis tools and their interconnectivities to form realistic
    multidisciplinary problems that is the subject of this paper. We are motivated by the following notional
    but real problem of organizing an MDO study for a complex system:

    A new system is being designed for which there is little or no historical precedent. The system
    is complex, as measured by the number of disciplines and/or components involved in the
    analysis. A general optimization problem statement has been formulated based on system-
    level objectives and constraints; however, it is unclear which engineering analysis tools should
    be interconnected in order to solve the optimization problem. A team of disciplinary and/
    or component design engineers has been formed in which each engineer has expertise in a
    particular analysis tool or component model. The engineers meet to discuss the approach to
    interconnecting their tools to achieve the required system-level MDO model.

    Our goal is to develop formalism for expressing analysis interconnectivity and for determining feasible
    data flows to assist an engineering team conducting this task. Because the problem deals with
    interconnectivity, we base our approach on the representations and techniques of graph theory. The 
    proposed graph syntax is closely related to the REMS system proposed by 
    Alexandrov and Lewis\cite{alexandrov2004}. The approach taken here should be viewed as an extention 
    of the concepts proposed in REMS with the goal of allowing the graphs to more closely interact with 
    existing design processes and MDAO frameworks. The approach begins by 
    constructing the \emph{maximal connectivity graph (MCG)} describing all possible
    interconnections between the analysis tools proposed by the engineers. Graph operations are then
    conducted reduce the MCG down to a \emph{fundamental problem graphs (FPG)} that describes the minimum set of analysis
    tools needed to solve the specified system-level design problem. The FPG does not predispose
    any particular solution procedure for the problem; any relevant MDO solution architecture, e.g. MDF,
    IDF, CO, BLISS, could be selected \emph{post facto} to implement the system-level optimization by
    constructing a \emph{problem solution graph (PSG)} and the corresponding problem solution approach.
    The concept of the FPG and the identification of feasible FPGs are the main contributions of the paper.

    The paper is organized as follows. First, we describe the differences between a fundamental problem
    formulation, which is based only on the system-level optimization problem statement that the
    engineers desire to solve and on the available analysis tools, and a specific problem formulation, which
    additionally presumes a specific solution to the problem. Next, we survey the literature in applying
    graph theoretic and formal language approaches to multidisciplinary design problem formulation. 
    We then discuss our graph syntax and representation of MDO problems and describe the procedures for 
    determining the MCG and FPG. Finally, we present an example problem based on an MDO analysis of a 
    commercial aircraft and discuss additional applications enabled by our approach.

@@

    The FPF in Eq. \ref{eqn:simple_fpf} differs from both Eq. \ref{eqn:simple} and Eq. \ref{eqn:simple2} because it has 
    two constraints which both must be met. The presence of both of these constraints fully decouples the problem so that 
    either $A$ or $B$ could be run first or both could be run simultaneously. Alexandrov and Lewis investigated problem 
    formulation in this fully decoupled way as part of their work on 
    Reconfigurability in MDO Problem Synthesis (REMS) \cite{alexandrov2004}.
    In this work they show how removing either constraint and replacing 
    it with a direct dependence between the two analyses, a process they term ``collapsing'', 
    you could regain the earlier two problem formulations.

\section{Existing Graph-Based Syntax}

    As shown above, the mathematical language for specifying problem formulations is very general and can be used both for 
    fundamental and specific problem formulations. Tedford and Martins used the above mathematical syntax to specify the 
    FPF for a set of test problems and also to describe specific formulations for solving them with a 
    number of optimization architectures\cite{Tedford2009}. Their work demonstrates clearly how multiple specific 
    problem formulations can all relate back to a common FPG. The challenge with using this 
    traditional mathematical syntax is that it is not easily manipulated or analyzed. 
    A number of graph-based methods have been used successfully to translate the 
    mathematical syntax into a more useful computational form.

@@

    Alexandrov and Lewis introduced a graph based syntax called REMS which 
    incorporates objectives and constraints into a graph, effectively combining 
    FDT and DSM\cite{alexandrov2004}. REMS retains the square adjacency 
    matrix form from DSM, but by adding in the new elements it partially 
    combined a traditional DSM with an FDT. This allows REMS to represent data 
    dependency between multiple analysis tools as well as between analysis tools and
    objective/constraint functions. One significant conclusion from REMS is 
    the need for variables that are local to discipline analyses. This allows 
    all information exchange to be stated as equality constraints, which become 
    edges in the graph. Additionaly, REMS address the need to be able to 
    transition between multiple solution strategies while keeping a constant 
    graph representation of the fundamental problem formulation. REMS only provides
    syntax for variables and functions, but does not provide for inclusion 
    of solvers or optimizers in the graph.

    Lamb and Martins also included objectives and constraint functions as nodes in an Extended 
    DSM (XDSM) in order to capture a more complete description of solution strategies for MDAO problems\cite{Lambe2012}. 
    Unlike REMS, XDSM also includes nodes for solvers and optimziers to enable complete definition of 
    MDAO architectures. Martins uses XDSM to describe 13 different optimziation architectures in 
    a survey paper that provides a novel classification of the different 
    techniques \cite{Lambe2011}. With the additional information included in an 
    XDSM, Lu and Martins applied both ordering and partitioning algorithms on an 
    MDAO test problem named the Scalable Problem \cite{Lu2012}.

@@

    Of all the methods, REMS comes the closest to fully describing a problem formulation,
    but lacks the ability to represent optimziers and solvers in more specific problem 
    formulations. While XDSM does include optimizers and constraints, it relies on 
    their presence and can't be used for a more fundamental problem statement. 
    We propose a new graph syntax that combines the key features of REMS and XDSM
    and provides new language to enable more complete description of MDAO problems.
